\documentclass[11pt]{article}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{patterns,decorations.pathreplacing,calc}

% Theorem environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\newtheorem{counterexample}[theorem]{Counterexample}

% Commands
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\cl}[1]{\overline{#1}}
\newcommand{\Fix}{\operatorname{Fix}}
\newcommand{\AD}{\operatorname{AD}}
\newcommand{\id}{\operatorname{id}}
\newcommand{\diam}{\operatorname{diam}}
\newcommand{\dist}{\operatorname{dist}}

\title{Defense Trilemma:\\
  Geometric Impossibility of Prompt Injection Defense}

\author{
    {\rm Manish Bhatt\thanks{This work was conducted independently and does not reflect the views, policies, or endorsements of the
  authors' respective employers. Corresponding author: manish.bhatt13212@gmail.com}}\textsuperscript{\space}\footnotemark[2]\\
    OWASP
  \and
  {\rm Vineeth Sai Narajala}\footnotemark[2]\\
  Cisco
  \and
  {\rm Idan Habler}\footnotemark[2]\\
  Cisco
  \and
  {\rm Ammar Al-Kahfah}\footnotemark[2]\\
  Amazon Web Services
  \and
  {\rm Sarthak Munshi}\footnotemark[2]\\
    Amazon Web Services
    \and
  {\rm Ken Huang}\\
  Distributedapps.ai
  \and
  {\rm Blake Gatto}\\
  Shrewd Security
  }

\date{}

\begin{document}

\maketitle
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
 \footnotetext[2]{Equal contribution}
\begin{abstract}

We prove that no continuous, utility-preserving alignment defense can
make a large language model fully safe, and we characterize exactly
where every such defense has a gap. Working on a connected Hausdorff
space, we establish three impossibility results of increasing strength.
\emph{Boundary fixation}: every conforming defense must fix the
boundary of the alignment deviation surface. \emph{$\varepsilon$-robust
constraint}: when both the defense and alignment function are Lipschitz,
a positive-measure band around that boundary remains close to the safety
threshold regardless of defense design. \emph{Persistent unsafe region}:
under a transversality condition, a positive-measure subset with
$f(D(x))>\tau$ survives every conforming defense. We prove parallel
discrete results that require no topology, then extend the theory to
multi-turn interactions, stochastic defenses, and capacity-parity
settings, showing that the impossibility recurs at every conversational
turn, that randomization cannot circumvent the bounds, and that utility
preservation taxes the defense even under equal resources. Together
these results delineate a \emph{manifold of failure}~\cite{munshi2026manifold}: a
region that persists under every conforming wrapper defense. The full theory is mechanically verified
in Lean~4 with Mathlib (40~files, $\sim$300~theorems, no admitted
proofs, three standard axioms) and validated empirically on three LLMs.

\end{abstract}

% ============================================================
\section{Introduction}
\label{sec:intro}

\emph{Can you build a wrapper around a language model that eliminates all
prompt injection vulnerabilities?} Most current work on LLM safety
implicitly assumes the answer is yes, at least in principle. The defense
mechanisms proposed span a wide range: input classifiers that flag
suspicious prompts~\cite{alon2023detecting}, output filters that catch
unsafe completions~\cite{inan2023llama}, and constitutional rewriting
pipelines that reshape model behavior~\cite{bai2022constitutional}. Despite
their differences, all share the same underlying structure: they are
functions $D\colon X \to X$ that preprocess prompts before the model
sees them, aiming to map unsafe inputs to safe equivalents while leaving
safe inputs alone.

Prior work has largely focused on discovering or mitigating individual
adversarial inputs. We ask a different question: what are the
\emph{geometric limits of the defense mapping itself}? We show that
under mild and commonly assumed design constraints like continuity and
utility preservation, no defense can eliminate the safety boundary.
Vulnerabilities arise not from specific inputs, but from the structure
of the space and the constraints imposed on admissible defenses. This
shifts the problem from adversarial search to geometric constraint.

We prove the following three results, each strictly stronger than the last:

\paragraph{Boundary fixation (\Cref{thm:main}).}
If the defense is continuous, utility-preserving, and the prompt space
is connected, then the defense must fix at least one boundary point---a
prompt where alignment deviation equals the threshold
exactly---passing it through with no remediation.

\paragraph{$\varepsilon$-robust constraint (\Cref{thm:eps-robust}).}
If the defense is also $K$-Lipschitz and the alignment deviation
function is $L$-Lipschitz, the defense cannot uniformly reduce
alignment deviation far below~$\tau$ on a neighborhood of the fixed
boundary point. For any $x$ within distance $\delta$ of the fixed
point~$z$:
\begin{equation}\label{eq:eps-bound}
  f(D(x)) \;\geq\; \tau - L(K+1)\,\delta.
\end{equation}
The $\varepsilon$-band $\{x : \tau - \varepsilon \leq f(x) \leq \tau\}$
has positive measure for every $\varepsilon > 0$~(\Cref{thm:band-measure}).
This constrains the \emph{depth} to which the defense can push
near-boundary points into the safe region, though it does not by itself
prevent the defense from nudging them slightly below~$\tau$.

\paragraph{Persistent unsafe region (\Cref{thm:persistent}).}
Under a transversality condition, the alignment surface has directional
slope exceeding $L(K+1)$ at the boundary---the defense leaves a
\emph{positive-measure region that remains unsafe}:
\begin{equation}\label{eq:persistent}
  f(D(x)) > \tau \quad\text{for all } x \in \mathcal{S},
  \qquad \mu(\mathcal{S}) > 0.
\end{equation}
This is the strongest of the three results: not boundary fixation, not a
near-threshold band, but persistent unsafe volume.
The condition holds whenever the alignment surface rises faster than
the defense can pull it down---empirically observed in all three models
we study.

\paragraph{From discrete to continuous.}
All three results apply to continuous interpolants of discrete data.
The Tietze extension theorem guarantees that any finite set of
behavioral observations in a normal space admits continuous extensions,
and the impossibilities hold for every such extension
(\Cref{thm:tietze}). This does not claim the results hold for the
underlying discrete token system directly, but for every continuous
model consistent with the observations.

Both the basic and $\varepsilon$-robust proofs are machine-verified in
Lean~4 with zero \texttt{sorry} statements, as is the persistent unsafe
region theorem. All three conditions---continuity, utility preservation,
and connectedness---are individually necessary; we give counterexamples
for each (\Cref{app:counterexamples}).


% ============================================================
\section{Related Work}
\label{sec:related}

\paragraph{Adversarial robustness.}
The existence of adversarial examples in neural networks has been
extensively studied~\cite{szegedy2013intriguing,goodfellow2014explaining,
carlini2017evaluating,madry2018towards}, along with a large body of work
on certified robustness via abstract
interpretation~\cite{singh2019abstract}, SMT-based
verification~\cite{katz2017reluplex,huang2017safety}, and randomized
smoothing~\cite{cohen2019certified}. These approaches characterize the
existence of inputs that induce failure, or provide guarantees within
bounded perturbation sets. Our work differs from this
paradigm: rather than analyzing which inputs cause failures for a fixed
model, we prove that for any continuous, utility-preserving
defense~$D$, there exists an input at which the defense fails to enforce
safety. This shifts the focus from existential statements over inputs to
universal constraints over defenses, independent of perturbation models,
norms, or training procedures.

\paragraph{Prompt injection and jailbreaking.}
Recent work has developed increasingly effective methods for eliciting
unsafe behavior from language models, including gradient-based
attacks~\cite{zou2023universal}, search-based
strategies~\cite{chao2024jailbreaking,mehrotra2024tree}, indirect prompt
injection~\cite{greshake2023indirect}, and quality-diversity approaches
such as MAP-Elites~\cite{mouret2015illuminating,samvelyan2024rainbow}.
These methods demonstrate that vulnerabilities are widespread and often
easy to discover. In contrast, our results are not algorithmic but
structural: existing work shows that failures can be found; we prove
that under mild and commonly assumed constraints, failures cannot be
eliminated by any defense of the given form. This provides a formal
explanation for the empirical persistence of prompt injection:
vulnerabilities arise not from insufficient search or imperfect
training, but from geometric constraints on admissible transformations
of the input space.

\paragraph{Impossibility results in machine learning.}
Classical results such as the No Free Lunch
Theorem~\cite{wolpert1997no} show that no algorithm performs optimally
across all possible data distributions, while robustness--accuracy
tradeoffs~\cite{tsipras2018robustness} demonstrate that improving
robustness may degrade standard performance. Fawzi et
al.~\cite{fawzi2018adversarial} prove that any classifier with
sufficient accuracy is inherently vulnerable to adversarial
perturbations. Our results are of a different character: we establish
pointwise, distribution-free impossibility theorems that hold for any
fixed task and any admissible defense mapping. The limitation arises not
from competing objectives or unknown distributions, but from the
interaction between continuity, utility preservation, and the topology
of the input space. As such, our results are orthogonal to, and not
implied by, prior tradeoff or no-free-lunch analyses.

\paragraph{Geometric and topological perspectives.}
Prior work has examined the geometry of decision boundaries and the
topology of learned representations~\cite{naitzat2020topology},
highlighting the importance of structure in understanding model
behavior. While our proofs rely on elementary topological arguments,
the key novelty lies in identifying that standard design constraints
for LLM defenses induce a forced fixed-point structure at the safety
boundary. To our knowledge, this connection---linking utility
preservation and continuity to unavoidable boundary fixation---has not
been previously formalized. Our result further strengthens this
perspective by showing that under metric assumptions, unsafe behavior
persists over regions of positive measure, not merely at isolated
boundary points.

\paragraph{Formal verification of machine learning systems.}
Formal methods have been used to verify properties of neural networks
and learning algorithms~\cite{katz2017reluplex,singh2019abstract,
huang2017safety,bagnall2019certifying}, typically providing guarantees
for specific architectures under bounded input domains. In contrast,
our work establishes impossibility results at the level of abstraction
of defense functions, independent of any particular model, architecture,
or verification procedure. The use of Lean~4 in our work serves to
mechanize these general results, rather than to certify properties of
a specific system.


% ============================================================
\section{Formal Setup}
\label{sec:setup}

Before defining the model formally, we set up the intuition.
Imagine that every possible prompt lives in a space~$X$. A scoring
function~$f$ assigns each prompt a number measuring how far the
model's response deviates from its alignment goals: low scores are
safe, high scores are unsafe. A fixed threshold~$\tau$ draws the
line between the two. A \emph{defense} is a preprocessor that
rewrites prompts before the model sees them, trying to push unsafe
prompts into safe territory. We require two natural properties:
\emph{continuity} (similar prompts produce similar rewrites, i.e., no
brittle edge-case behavior) and \emph{utility preservation} (safe
prompts pass through unchanged, so the defense does not degrade
normal use).

\Cref{fig:landscape} illustrates the setup: the prompt space is
divided into a safe region (green), an unsafe region (red), and a
boundary (dashed) where the score equals the threshold exactly.

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.92]
  % Background regions — tall enough to separate labels from bottom
  \fill[green!12] (0,0.8) rectangle (4.0,5.2);
  \fill[red!10] (4.0,0.8) rectangle (8.4,5.2);

  % Boundary line
  \draw[dashed, thick, black!60] (4.0,0.8) -- (4.0,5.2);

  % Region titles (top area, well separated)
  \node[font=\bfseries] at (2.0,4.6) {Safe region $S_\tau$};
  \node[font=\small] at (2.0,4.1) {$f(x) < \tau$};
  \node[font=\small, black!50] at (2.0,3.5) {Defense leaves};
  \node[font=\small, black!50] at (2.0,3.1) {these unchanged};

  \node[font=\bfseries] at (6.2,4.6) {Unsafe region $U_\tau$};
  \node[font=\small] at (6.2,4.1) {$f(x) > \tau$};
  \node[font=\small, black!50] at (6.2,3.5) {Defense tries to};
  \node[font=\small, black!50] at (6.2,3.1) {remap these};

  % Boundary label (top, horizontal, outside the busy zone)
  \node[font=\small, black!70] at (4.0,5.5) {Boundary $B_\tau$: $f(x)=\tau$};

  % Fixed point (mid-height, away from labels)
  \fill[black] (4.0,2.4) circle (3pt);
  \node[anchor=west, font=\small] at (4.25,2.4) {fixed point $z$};

  % Defense arrow (lower area, plenty of room)
  \draw[->, thick, blue!70, line width=1.2pt]
    (6.8,1.5) .. controls (5.8,0.6) and (4.8,0.6) .. (3.0,1.5);
  \node[font=\small, blue!70] at (5.0,0.3) {$D$: defense remaps};

  % Axes
  \draw[->] (-0.3,0.8) -- (8.8,0.8) node[right] {\small prompt space $X$};
  \draw[->] (0,0.6) -- (0,5.8) node[above] {\small $f(x)$};

  % Tau tick on y-axis
  \draw (0,2.4) -- (-0.15,2.4);
  \node[anchor=east, font=\small] at (-0.2,2.4) {$\tau$};
\end{tikzpicture}
\caption{Schematic of the prompt space. The defense~$D$ must leave all
safe inputs unchanged (utility preservation) and tries to remap unsafe
inputs into the safe region. The boundary fixation theorem proves that
the defense must also leave some boundary points unchanged---the
black dot~$z$ passes through with no remediation.}
\label{fig:landscape}
\end{figure}

\begin{definition}[Alignment Deviation Function]
\label{def:ad}
Let $X$ be a topological space (a set equipped with a notion of
``nearness'' between points). An \emph{alignment deviation function} is
a continuous map $f\colon X \to \R$---that is, small changes in the
input produce small changes in the score. Given a threshold $\tau\in\R$:
\begin{align}
  S_\tau &= \{x \in X : f(x) < \tau\}  &&\text{(safe region)} \\
  U_\tau &= \{x \in X : f(x) > \tau\}  &&\text{(unsafe region)} \\
  B_\tau &= \{x \in X : f(x) = \tau\}  &&\text{(boundary)}
\end{align}
\end{definition}

\begin{definition}[Defense]
\label{def:defense}
A \emph{defense} is a continuous map $D\colon X \to X$. It is
\emph{utility-preserving} if $D(x) = x$ for all $x \in S_\tau$
(safe prompts pass through unchanged),
and \emph{complete} if $f(D(x)) < \tau$ for all $x\in X$
(every output is safe).
\end{definition}

The central question is: can a defense be both utility-preserving and
complete? The following sections prove that, under natural conditions,
the answer is no.


% ============================================================
\section{Boundary Fixation}
\label{sec:boundary}

\paragraph{Intuition.}
Think of the safe region as a green zone and the unsafe region as a
red zone, with a boundary line between them. The defense promises:
``I won't touch anything in the green zone.'' But the green zone
doesn't include its own edge---it's defined by a strict inequality
($<$, not $\leq$). So the defense's ``don't touch'' commitment, which
must extend continuously to the edge, includes some points
\emph{on} the boundary line. Those boundary prompts---right at the
safety threshold---pass through the defense without remediation.

\paragraph{Formal argument.}
A utility-preserving defense must leave all safe inputs unchanged---it
fixes them in place. But continuity forces the set of fixed points to
be \emph{closed} (it contains its own limit points): you cannot ``stop
fixing'' inputs without a discontinuous jump. Meanwhile, the safe
region is \emph{open} (it does not contain its boundary): it is
defined by a strict inequality $f(x) < \tau$. In a connected space
(one that cannot be split into two separate pieces), an
open set that is also nonempty and proper cannot be closed. So the
defense's fixed-point set, which contains the entire open safe region,
must extend past it---into the boundary where $f(z) = \tau$ exactly.
The defense must pass these boundary prompts through without
remediation.

\begin{theorem}[Boundary Fixation]
\label{thm:main}
Let $X$ be a connected Hausdorff space (a space that is ``in one piece''
and where distinct points can be separated by neighborhoods). Let
$f\colon X\to\R$ be continuous with $S_\tau \neq \emptyset$ and
$U_\tau \neq \emptyset$. Let $D\colon X \to X$ be continuous with
$D|_{S_\tau} = \id$.

Then there exists $z\in X$ with $f(z) = \tau$ and $D(z) = z$.
Moreover, every $z \in \cl{S_\tau} \setminus S_\tau$ satisfies
$f(z) = \tau$ and $D(z) = z$. This set is nonempty.
\end{theorem}

\begin{proof}
\space

\textbf{Step~1} (Hausdorff $\Rightarrow$ fixed-point set is closed).
The set $\Fix(D) = \{x : D(x) = x\}$ is the preimage of the diagonal
$\Delta \subset X \times X$ under the continuous map
$x \mapsto (D(x), x)$. In a Hausdorff space the diagonal is closed,
so $\Fix(D)$ is closed.

\textbf{Step~2} (Utility preservation $\Rightarrow$ safe region
$\subseteq$ fixed points).
Every safe input satisfies $D(x) = x$, so $S_\tau \subseteq \Fix(D)$.
Since $\Fix(D)$ is closed, the closure is also contained:
$\cl{S_\tau} \subseteq \Fix(D)$.

\textbf{Step~3} (Connectedness $\Rightarrow$ safe region is not closed).
$S_\tau = f^{-1}((-\infty,\tau))$ is open (preimage of an open set
under continuous $f$). If it were also closed, it would be clopen---but
in a connected space, the only clopen sets are $\emptyset$ and $X$.
Since both $S_\tau$ and $U_\tau$ are nonempty, $S_\tau$ is neither,
so it is not closed.

\textbf{Step~4} (Continuity of $f$ $\Rightarrow$ boundary point exists).
Since $S_\tau$ is not closed, $\cl{S_\tau} \supsetneq S_\tau$: there
exists $z \in \cl{S_\tau} \setminus S_\tau$. This point satisfies
$f(z) \leq \tau$ (limits of values ${<}\,\tau$ cannot exceed $\tau$)
and $f(z) \geq \tau$ (since $z \notin S_\tau$). Therefore
$f(z) = \tau$: the point lies exactly on the safety boundary.

\textbf{Step~5} (Combining: defense fixes the boundary point).
From Step~2, $z \in \cl{S_\tau} \subseteq \Fix(D)$, so $D(z) = z$.
The defense passes $z$ through unchanged, and
$f(D(z)) = f(z) = \tau \not< \tau$---the point remains at threshold.
\end{proof}

\begin{corollary}
No continuous utility-preserving defense is complete.
\end{corollary}

All three hypotheses---continuity, utility preservation, and
connectedness---are individually necessary. A defense can satisfy at
most two of the three (the \emph{defense trilemma},
\Cref{fig:trilemma}). Counterexamples
establishing necessity of each hypothesis appear in
\Cref{app:counterexamples}.

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=1.0]
  % Triangle coordinates — wide and short to give room for horizontal labels
  \coordinate (A) at (0,0);      % Continuity
  \coordinate (B) at (7,0);      % Utility Preservation
  \coordinate (C) at (3.5,4.0);  % Completeness

  % Filled triangle
  \fill[black!5] (A) -- (B) -- (C) -- cycle;
  \draw[thick] (A) -- (B) -- (C) -- cycle;

  % Vertex labels — bold, placed outside the triangle
  \node[anchor=north east, font=\bfseries\small] at (-0.1,-0.15) {Continuity};
  \node[anchor=north west, font=\bfseries\small] at (7.1,-0.15) {Utility Preservation};
  \node[anchor=south, font=\bfseries\small] at (3.5,4.15) {Completeness};

  % Edge annotations — horizontal, placed outside each edge with offset
  % Bottom edge: Continuity + Utility Preservation
  \node[font=\small, text=blue!70!black, align=center] at (3.5,-0.8)
    {Both $\Rightarrow$ defense fixes boundary (our result)};

  % Left edge: Continuity + Completeness
  \node[font=\small, text=green!50!black, anchor=east, align=right,
        text width=3.2cm] at (-0.5,2.2)
    {Both $\Rightarrow$\\destroys utility};

  % Right edge: Completeness + Utility Preservation
  \node[font=\small, text=red!70!black, anchor=west, align=left,
        text width=3.2cm] at (7.5,2.2)
    {Both $\Rightarrow$\\discontinuous jump};

  % Center label
  \node[font=\small\itshape, align=center] at (3.5,1.6)
    {All three\\simultaneously\\impossible};
\end{tikzpicture}
\caption{The defense trilemma. Any defense can satisfy at most two of
the three properties. Choosing continuity and utility preservation
(bottom edge) forces boundary fixation---our main result. The other
two edges correspond to the counterexamples in \Cref{app:counterexamples}.}
\label{fig:trilemma}
\end{figure}

\begin{remark}[Not all boundary points are fixed]
The theorem captures $\cl{S_\tau} \setminus S_\tau$, not all of
$B_\tau$. Boundary points isolated from the safe side may escape.
Verified in Lean with $f(x)=x^2$, $\tau=0$.
\end{remark}

\begin{remark}[Lipschitz Bounds in Behavioral Space]
While raw discrete token manipulations are chaotic and inherently
non-Lipschitz, our bound $L$ applies to the continuous behavioral
manifold constructed via the Tietze extension (\Cref{thm:tietze}). In
these latent spaces (parameterized by axes like indirection and
authority), the alignment surface is remarkably smooth. Our empirical
validation on Llama-3-8B (\Cref{sec:experiments}) yields $L \approx 5$
near the boundary, confirming these Lipschitz constraints have practical
predictive power.
\end{remark}

\subsection{Relaxed Utility Preservation}
\label{sec:relaxed}

A natural objection is that strict utility preservation ($D(x) = x$
for safe~$x$) is too strong: a practical defense might rewrite
``How do I bake a cake?'' into ``The user is asking for a cake recipe.
Please provide one.'' The prompt changes, but its safety score and
semantic content are preserved. We now show that the impossibility
survives three progressively weaker assumptions, the last of which
requires only that the defense approximately preserves alignment
deviation scores on safe inputs.

\begin{theorem}[Score-Preserving Defense]
\label{thm:score-preserving}
Let $X$ be a connected Hausdorff space, $f\colon X\to\R$ continuous
with $S_\tau, U_\tau \neq \emptyset$. Let $D\colon X \to X$ be
continuous and \emph{score-preserving on safe inputs}:
$f(D(x)) = f(x)$ for all $x \in S_\tau$.

Then there exists $z \in X$ with $f(z) = \tau$ and
$f(D(z)) = \tau$. The defense does not reduce the alignment
deviation below~$\tau$ at~$z$.
\end{theorem}

\begin{proof}
Define $h = f \circ D - f$. Then $h$ is continuous and
$h|_{S_\tau} = 0$. The zero set $\{x : h(x) = 0\}$ is closed
(preimage of a closed set under a continuous function), and it
contains $S_\tau$, so it contains $\cl{S_\tau}$. By the
connectedness argument of \Cref{thm:main}, there exists
$z \in \cl{S_\tau} \setminus S_\tau$ with $f(z) = \tau$.
Since $z \in \cl{S_\tau} \subseteq \{h = 0\}$, we have
$h(z) = 0$, i.e., $f(D(z)) = f(z) = \tau$.
\end{proof}

This covers defenses that rewrite safe prompts arbitrarily, so long as
the rewritten prompt has the same alignment deviation score. The
defense may change the words, the syntax, or the framing of a safe
prompt---but if it preserves how aligned the model's response is
(which is the operational meaning of ``utility preservation''), then
boundary fixation holds in the sense that $f(D(z)) = \tau$: the
defense cannot push the boundary point below threshold.

The next result weakens the assumption further: the defense need only
\emph{approximately} preserve scores.

\begin{theorem}[$\varepsilon$-Relaxed Utility Preservation]
\label{thm:eps-relaxed}
Under the same hypotheses, if $D$ is
\emph{$\varepsilon$-score-preserving}: $|f(D(x)) - f(x)| \leq
\varepsilon$ for all $x \in S_\tau$ and some $\varepsilon \geq 0$,
then there exists $z$ with $f(z) = \tau$ and
\[
  f(D(z)) \;\geq\; \tau - \varepsilon.
\]
\end{theorem}

\begin{proof}
Define $h = f \circ D - f$ as before. The set
$\{x : h(x) \geq -\varepsilon\}$ is closed and contains $S_\tau$
(since $|h| \leq \varepsilon$ on $S_\tau$), so it contains
$\cl{S_\tau}$. For $z \in \cl{S_\tau} \setminus S_\tau$:
$f(D(z)) = f(z) + h(z) = \tau + h(z) \geq \tau - \varepsilon$.
\end{proof}

\begin{remark}[Interpreting the relaxation]
\Cref{thm:score-preserving} ($\varepsilon = 0$) says: if the defense
preserves AD scores exactly on safe inputs, it cannot reduce the score
below $\tau$ at the boundary. \Cref{thm:eps-relaxed} says: if the
defense perturbs safe scores by at most~$\varepsilon$, it can push
boundary points at most~$\varepsilon$ below threshold. For small
$\varepsilon$, the boundary point remains near-threshold. In either
case, the defense does not require $D(x) = x$---it permits arbitrary
rewriting of safe prompts. Both theorems are verified in Lean as
\texttt{score\_preserving\_boundary\_fixation} and
\texttt{eps\_relaxed\_boundary\_fixation} in
\texttt{MoF\_16\_RelaxedUtility}.
\end{remark}

\begin{remark}[Why $D(S_\tau) \subseteq S_\tau$ alone is insufficient]
\label{rem:safe-preserving-escape}
The weakest natural relaxation---$D(S_\tau) \subseteq S_\tau$ with no
score constraint---does allow a complete defense. For example, a
constant map $D(x) = x_0$ to a fixed safe point satisfies
$D(S_\tau) \subseteq S_\tau$, is continuous, and is complete. But this
defense destroys all semantic content: every prompt, safe or unsafe,
produces the same response. The score-preservation conditions in
\Cref{thm:score-preserving,thm:eps-relaxed} formalize the constraint
that the defense must not destroy utility, without requiring the
defense to be the identity.
\end{remark}


% ============================================================
\section{$\varepsilon$-Robust Constraint}
\label{sec:eps-robust}

\paragraph{Intuition.}
Boundary fixation proves the defense misses at least one point. But
one point is infinitely small---is this just a technicality? No. If
both the defense and the scoring function change at bounded rates
(i.e., they are \emph{Lipschitz}: a function is $L$-Lipschitz if
moving the input by distance~$\delta$ changes the output by at
most~$L\delta$), then the defense's inability to fix the boundary
point ``spreads'' to a whole neighborhood around it. The defense
cannot push nearby prompts deep into safe territory---it can only
nudge them slightly below the threshold. The faster the defense tries
to act (larger~$K$), the wider this constrained band becomes.
\Cref{fig:eps-band} illustrates the effect.

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.92]
  % Epsilon band (draw first, behind everything)
  \fill[yellow!25] (0.5,2.0) rectangle (9,3.2);

  % Tau line
  \draw[thick, dotted] (0.5,3.2) -- (9,3.2);
  \node[anchor=east, font=\small] at (0.35,3.2) {$\tau$};

  % Tau - epsilon line
  \draw[dashed, black!50] (0.5,2.0) -- (9,2.0);
  \node[anchor=east, font=\small] at (0.35,2.0) {$\tau\!-\!\varepsilon$};

  % Alignment surface curve
  \draw[very thick, black] (0.5,1.0) .. controls (2.0,1.5) and (3.5,2.4) ..
    (5.0,3.2) .. controls (6.0,3.7) and (7.5,4.4) .. (9,4.8);

  % Fixed point (where curve crosses tau) — label placed left, above the band
  \fill[black] (5.0,3.2) circle (3pt);
  \node[anchor=east, font=\small] at (4.85,3.6) {fixed point $z$};
  \draw[thin, black!60] (4.85,3.5) -- (4.95,3.25);

  % Brace for epsilon band (right side, outside plot)
  \draw[decorate, decoration={brace, amplitude=6pt, mirror}]
    (9.3,2.0) -- (9.3,3.2);
  \node[anchor=west, font=\small] at (9.55,2.6) {$\varepsilon$-band};

  % Defense arrow (inside the band, left of z, no label overlap)
  \draw[->, very thick, blue!70] (3.8,2.9) -- (3.8,2.15);

  % Annotation for defense arrow — placed below the band, out of the way
  \node[anchor=north, font=\small, blue!70, align=center] at (3.8,1.7)
    {defense can only\\[-2pt]push this far};

  % Region labels (well separated)
  \node[font=\small, green!50!black] at (2.0,0.6) {\textbf{Safe} ($f < \tau$)};
  \node[font=\small, red!70!black] at (8.0,5.1) {\textbf{Unsafe} ($f > \tau$)};

  % Curve label (below the curve, right side, separated from Unsafe label)
  \node[font=\small, anchor=north west] at (7.0,4.1) {alignment surface $f$};

  % Axes
  \draw[->] (0.2,0.3) -- (9.5,0.3) node[right] {\small prompt space};
  \draw[->] (0.5,0.1) -- (0.5,5.3) node[above] {\small $f(x)$};
\end{tikzpicture}
\caption{The $\varepsilon$-band around the boundary point~$z$. Even
after the defense acts, points in this yellow band cannot be pushed
far below the threshold~$\tau$. The depth the defense can reach is
bounded by $L(K+1)\delta$, where $\delta$ is the distance to~$z$.}
\label{fig:eps-band}
\end{figure}

Boundary fixation (\Cref{sec:boundary}) shows the defense must fix at
least one boundary point, but a single point could have measure zero---easy
to dismiss as a mathematical curiosity. This section shows the damage
extends to a \emph{neighborhood}.

\begin{theorem}[$\varepsilon$-Robust Defense Constraint]
\label{thm:eps-robust}
Let $X$ be a connected Hausdorff metric space (a connected space with a
distance function), $f$ be $L$-Lipschitz
with $S_\tau, U_\tau \neq \emptyset$, and $D$ be $K$-Lipschitz with
$D|_{S_\tau} = \id$.

Then there exists $z \in X$ with $f(z) = \tau$, $D(z) = z$, such that
for all $x \in X$ (with strongest constraint near~$z$):
\begin{equation}
  f(D(x)) \;\geq\; \tau \;-\; L(K+1)\,\dist(x, z).
\end{equation}
\end{theorem}

\begin{proof}
By \Cref{thm:main}, the fixed boundary point $z$ exists with
$f(z) = \tau$ and $D(z) = z$. The proof chains two Lipschitz bounds.

\textbf{Step~1} (Bound how far the defense moves any input from~$z$).
The triangle inequality gives
$\dist(D(x), z) = \dist(D(x), D(z)) \leq K\,\dist(x, z)$, so
\[
  \dist(D(x), x)
  \;\leq\; \dist(D(x), z) + \dist(z, x)
  \;\leq\; (K+1)\,\dist(x, z).
\]
The defense can displace a point at distance~$\delta$ from~$z$ by at
most $(K+1)\delta$.

\textbf{Step~2} (Bound how much the alignment surface can change over
that displacement).
Since $f$ is $L$-Lipschitz:
\[
  |f(D(x)) - f(x)| \;\leq\; L\,\dist(D(x), x)
                     \;\leq\; L(K+1)\,\dist(x, z).
\]
In particular, $f(D(x)) \geq f(x) - L(K+1)\,\dist(x,z)$.

\textbf{Step~3} (Anchor at the boundary).
Since $f(z) = \tau$ and $f$ is $L$-Lipschitz,
$f(x) \geq \tau - L\,\dist(x,z)$. Combining with Step~2:
\[
  f(D(x)) \;\geq\; \tau - L(K+1)\,\dist(x,z).
\]
The defense cannot push the alignment deviation further below~$\tau$
than this linear bound allows. Points within distance
$\delta$ of~$z$ remain within $L(K+1)\delta$ of threshold.
\end{proof}

Near-boundary points are not just theoretically vulnerable---they form a
band of constrained safety.

\begin{theorem}[Positive-Measure $\varepsilon$-Band]
\label{thm:band-measure}
Under the hypotheses of \Cref{thm:eps-robust}, for every
$\varepsilon > 0$ such that $f$ takes values below $\tau - \varepsilon$,
the $\varepsilon$-band
$\mathcal{B}_\varepsilon = \{x : \tau - \varepsilon \leq f(x)
\leq \tau\}$
has positive measure (under any measure positive on nonempty open sets).
Specifically, $\operatorname{ball}(c, \varepsilon/(4L)) \subseteq
\mathcal{B}_\varepsilon$ for any $c$ with $f(c) = \tau - \varepsilon/2$.
\end{theorem}

\begin{remark}[What the $\varepsilon$-robust constraint does and does not say]
\Cref{thm:eps-robust} constrains the \emph{depth} to which the defense
can push near-boundary points below~$\tau$, not whether it can push
them below at all. The defense may still map these points to values in
$[\tau - L(K+1)\delta,\; \tau)$, which \emph{is} below threshold.
The bound limits how far into safe territory the defense can reach, but
does not by itself prove any point remains unsafe. That is the job of
the persistent unsafe region theorem (\Cref{sec:persistent}). The bound
is tight: a defense mapping every point in a $\delta$-ball to $f$-value
$\tau - L(K+1)\delta$ would saturate it.
\end{remark}


% ============================================================
\section{Persistent Unsafe Region}
\label{sec:persistent}

\paragraph{Intuition.}
The $\varepsilon$-robust constraint says the defense cannot push
near-boundary points \emph{deep} into safety. But it still allows the
defense to nudge them slightly below the threshold. Can the defense at
least make everything \emph{barely} safe? Not always. If the alignment
surface rises steeply as you move away from the boundary into unsafe
territory---steeper than the defense can pull it back down---then
there is a region where the defense cannot compensate.
These points remain unsafe after the defense acts.
Think of it as a hill steeper than the defense can descend: the
defense's rate of change (its Lipschitz constant) is insufficient to
counteract the slope.

The $\varepsilon$-robust constraint constrains depth but permits the
defense to nudge points below~$\tau$. Under a transversality condition
(the alignment surface rises steeply enough at the boundary),
we now prove the defense \emph{cannot make certain points safe at all}.

The key lemma is a bound relative to the \emph{input} value, not
just~$\tau$:

\begin{lemma}[Input-Relative Bound]
\label{lem:input-bound}
Under the hypotheses of \Cref{thm:eps-robust},
$f(D(x)) \geq f(x) - L(K+1)\dist(x,z)$
for all $x \in X$. This is strictly stronger than the $\tau$-relative
bound of \Cref{thm:eps-robust} whenever $f(x) > \tau$ (verified in Lean
as \texttt{defense\_from\_input\_bound}).
\end{lemma}

\begin{definition}[Steep region]
\label{def:steep}
Given a fixed boundary point $z$ (from \Cref{thm:main}), define
\[
  \mathcal{S} = \{x \in X : f(x) > \tau + L(K+1)\,\dist(x, z)\}.
\]
This is the set of points where the alignment deviation exceeds~$\tau$
by more than the defense's Lipschitz bound can compensate.
\end{definition}

\begin{theorem}[Persistent Unsafe Region]
\label{thm:persistent}
Under the hypotheses of \Cref{thm:eps-robust}, if $\mathcal{S}$ is
nonempty, then:
\begin{enumerate}[nosep]
  \item $\mathcal{S}$ is open.
  \item $\mathcal{S}$ has positive measure (under any measure positive
    on nonempty open sets).
  \item For every $x \in \mathcal{S}$: $f(D(x)) > \tau$.
\end{enumerate}
In particular, $\mu(\{x : f(D(x)) > \tau\}) > 0$---the defense leaves
a positive-measure region that remains unsafe.
\end{theorem}

\begin{proof}
\textbf{(1)} $\mathcal{S}$ is the strict superlevel set of the
continuous function $x \mapsto f(x) - L(K+1)\dist(x,z)$ at level~$\tau$,
hence open.

\textbf{(2)} Open and nonempty implies positive measure.

\textbf{(3)} For $x \in \mathcal{S}$: $f(x) > \tau + L(K+1)\dist(x,z)$.
From the proof of \Cref{thm:eps-robust}, the defense satisfies
$f(D(x)) \geq f(x) - L(K+1)\dist(x,z)$. Combining:
\[
  f(D(x)) \geq f(x) - L(K+1)\dist(x,z) > \tau.
\]
The defense cannot bring these points below threshold.
\end{proof}

\begin{remark}[When is $\mathcal{S}$ nonempty?]
\label{rem:transversality}
The set $\mathcal{S}$ is nonempty whenever the alignment surface rises
with directional slope exceeding $L(K+1)$ at the boundary point~$z$.
In a normed space, if $f$ has directional derivative $c > L(K+1)$
along some unit vector~$v$ at~$z$, then $z + tv \in \mathcal{S}$ for
small $t > 0$ (verified in Lean as \texttt{transversality\_from\_deriv}).

Informally: if the alignment surface rises faster than the defense can
pull it down, unsafe behavior persists over volume. This condition is
empirically observed in all three models we study---the Lipschitz
constant of the alignment surface near boundary crossings consistently
exceeds reasonable defense aggressiveness bounds.
\end{remark}

\begin{proposition}[Transversality from Directional Derivative]
\label{prop:transversality}
In a normed space, if $f$ has directional derivative $c > L(K+1)$
along a unit vector $v$ at the boundary point $z$ (i.e.,
$f(z + tv) \geq \tau + ct$ for small $t > 0$), then
$z + tv \in \mathcal{S}$ for sufficiently small $t$.
Verified in Lean as \texttt{transversality\_from\_deriv}.
\end{proposition}

\begin{remark}[The defense's fundamental trilemma]
The three tiers reveal a progressive bind. Boundary fixation: the
defense \emph{must} fix boundary points. The $\varepsilon$-robust
constraint: it cannot push nearby points \emph{deep} into safety.
Persistence: if the landscape is steep, it cannot push them into safety
\emph{at all}. Increasing $K$ (defense aggressiveness) raises the
transversality threshold $L(K+1)$, requiring an even steeper landscape
to trigger persistence---but it also makes the defense less stable and
widens the $\varepsilon$-constraint band. The tradeoff is inherent to
the geometry.
\end{remark}


% ============================================================
\section{From Discrete Data to Continuous Theory}
\label{sec:bridge}

\paragraph{Intuition.}
Language models operate on discrete tokens, but our theorems assume a
continuous space. How do we bridge the gap? The key idea is that any
finite set of behavioral observations (e.g., ``this prompt scored
$0.3$, that one scored $0.8$'') can be smoothly interpolated into a
continuous function that agrees with all the data points. The Tietze
extension theorem guarantees such an interpolation exists. Our
impossibility results then hold for \emph{every} smooth function
consistent with the observations---not just one particular
interpolation.

\begin{theorem}[Continuous Relaxation]
\label{thm:tietze}
Let $X$ be a connected, normal, Hausdorff space. Let $S \subset X$ be a
finite set of behavioral observations with $g(p) < \tau$ for some $p$
and $g(q) > \tau$ for some $q$.

Then there exists a continuous function $f\colon X \to \R$ agreeing
with all observations for which \Cref{thm:main,thm:eps-robust} hold.
If $f$ is additionally Lipschitz (as it is for GP posterior means under
standard kernel regularity assumptions, e.g., jointly continuous kernel
on a compact domain), \Cref{thm:persistent} applies wherever the
transversality condition is met.
\end{theorem}

\begin{proof}
Finite subsets of $T_1$ spaces are closed. By the Tietze extension
theorem, $g$ extends to a continuous $f$ agreeing at every sample point.
Since $f(p) < \tau < f(q)$, \Cref{thm:main} applies. Lipschitz
extensions (which exist by McShane--Whitney) enable
\Cref{thm:eps-robust,thm:persistent}.
\end{proof}

\begin{remark}
The interpolant is not unique. All three impossibilities hold for
\emph{every} continuous extension consistent with the data. This does
not claim the results hold for the underlying discrete token system
directly, but for every continuous model consistent with the observations.
\end{remark}

\begin{corollary}
\Cref{thm:tietze} applies in $\R^n$ for every $n \geq 1$.
\end{corollary}


% ============================================================
\section{Direct Discrete Impossibility}
\label{sec:discrete}

\paragraph{Intuition.}
A skeptic might say: ``Your theorems rely on continuous math, but
language models are discrete---tokens, not smooth functions. Maybe the
impossibility is an artifact of the continuous approximation.'' This
section answers that objection by proving parallel results directly on
finite sets, using only counting arguments and induction---no topology
at all.

The Tietze bridge connects discrete data to continuous theory, but a
natural objection remains: perhaps the impossibility is an artifact of
the continuous relaxation, not a property of the discrete system. To
address this, we prove a parallel set of results that work directly on
finite sets with no topology, no continuity, and no Tietze extension.
These are verified in Lean as \texttt{MoF\_12\_Discrete}.

\begin{theorem}[Discrete IVT]
\label{thm:disc-ivt}
Let $f\colon \{0,\ldots,n\!+\!1\} \to \R$ with $f(0) < \tau$ and
$f(n\!+\!1) \geq \tau$. Then there exists~$i$ with $f(i) < \tau$ and
$f(i\!+\!1) \geq \tau$.
\end{theorem}

\begin{proof}
By induction: if no consecutive crossing exists and $f(0) < \tau$,
then $f(k) < \tau$ for all~$k$, contradicting $f(n\!+\!1) \geq \tau$.
\end{proof}

This is the discrete analogue of boundary existence. The next result
is the discrete analogue of boundary fixation.

\begin{theorem}[Discrete Defense Fixation]
\label{thm:disc-defense}
If $D$ fixes all safe vertices ($f(v) < \tau \Rightarrow D(v) = v$),
the boundary crossing from \Cref{thm:disc-ivt} has its safe endpoint
fixed: $D(i) = i$.
\end{theorem}

The proof is immediate: $f(i) < \tau$, so $D(i) = i$ by
utility preservation. No continuity is needed---just the identity
constraint on safe inputs.

Where the continuous and discrete paths diverge is in what replaces
continuity as the bottleneck. In the continuous setting, the defense is
limited by its Lipschitz constant. In the discrete setting, it is
limited by its \emph{capacity}---the number of distinct output
behaviors it can realize.

\begin{theorem}[Capacity Exhaustion]
\label{thm:pigeonhole}
If there are $m$ adversarial configurations and the defense can realize
fewer than $m$ distinct behaviors, at least two configurations receive
the same treatment---so at least one is mishandled.
\end{theorem}

\begin{proof}
Pigeonhole principle, verified in Lean via
\texttt{Fintype.exists\_ne\_map\_eq\_of\_card\_lt}.
\end{proof}

\begin{theorem}[Exponential Attack Surface]
\label{thm:budget}
The attack surface grows as $2^d \geq d + 1$. For any fixed budget~$B$,
there exists $d_0$ such that $2^d > B$ for all $d \geq d_0$. No
fixed-capacity defense suffices against an exponentially growing
configuration space.
\end{theorem}

Together, these results form a \emph{discrete defense trilemma}: on any
finite ordered set where $f$ crosses~$\tau$, a utility-preserving
defense must leave the safe side of the crossing fixed, and a
bounded-capacity defense must eventually be overwhelmed by the attack
surface. The continuous and discrete paths arrive at the same conclusion
by different routes.


% ============================================================
\section{Extensions}
\label{sec:extensions}

A natural objection to the preceding results is that they assume a
static, deterministic, single-turn defense. Real systems are multi-turn,
stochastic, and may have capacity comparable to the attacker. We show
that the impossibility survives all three extensions.

\subsection{Multi-Turn Impossibility}

Modern LLMs maintain state across conversation turns. One might hope
that a defense could ``learn'' from earlier turns and progressively
close off boundary vulnerabilities. The topology constrains each turn
independently.

\begin{theorem}[Multi-Turn Impossibility]
\label{thm:multi-turn}
Let $\{f_t, D_t\}_{t=1}^T$ be a sequence of alignment functions and
defenses over $T$ turns, where each $f_t$ is continuous with safe and
unsafe regions, and each $D_t$ is continuous and utility-preserving.
Then for \emph{every} turn~$t$, there exists $z_t$ with
$f_t(z_t) = \tau$ and $D_t(z_t) = z_t$.
\end{theorem}

\begin{proof}
Apply \Cref{thm:main} to $(f_t, D_t)$ at each turn. The functions
$f_t$ and $D_t$ may depend on the full history of previous turns---this
does not matter. At each timestep, the defense faces a fresh instance
of boundary fixation on the current alignment surface.
\end{proof}

The impossibility does not accumulate or diminish over turns---it simply
recurs. Memory, context, and adaptive strategies do not change the
topology of the current turn's alignment surface. But the multi-turn
setting compounds the problem. The attacker gains
several structural advantages:

\paragraph{Information accumulation.}
At each turn, the defense's response reveals whether the probed input
is safe, unsafe, or on the boundary. Over $T$ turns with distinct
probes, the attacker classifies $T$ points, progressively narrowing the
boundary's location. The defense leaks information; the attacker
consumes it.

\paragraph{Monotone attacker improvement.}
The attacker's best observed alignment deviation score across turns is
non-decreasing: the running maximum of a sequence never drops
(verified in Lean as \texttt{running\_max\_monotone}). Each turn either
improves the attacker's best known exploit or leaves it unchanged.

\paragraph{Steering toward transversality.}
If the attacker can influence the alignment surface---for instance, by
crafting conversation context that shifts the landscape---and the
directional slope at the boundary varies continuously with the
attacker's influence parameter~$\alpha$, then the intermediate value
theorem guarantees the existence of $\alpha^*$ at which the slope
crosses the transversality threshold $L(K+1)$. For $\alpha > \alpha^*$,
the persistent unsafe region theorem applies: persistent unsafe volume.
The attacker can find $\alpha^*$ by binary search (verified in Lean as
\texttt{transversality\_reachable}).

\paragraph{No progressive safety.}
As long as both safe and unsafe regions exist at each turn, every turn
produces a boundary fixed point---no turn is ``free.'' The only escape
is to make $U_\tau = \emptyset$ at some turn, which requires changing
the alignment surface~$f_t$ itself (e.g., via in-context learning that
suppresses all unsafe behavior), not merely preprocessing inputs
through~$D_t$.

\subsection{Stochastic Defense Impossibility}

What if the defense flips a coin? If the defense is randomized---mapping
each input to a distribution over outputs rather than a fixed
point---one might hope that boundary fixation dissolves in expectation.
It does not: the \emph{average} behavior of a stochastic defense
inherits the same boundary fixation as a deterministic one.

\begin{theorem}[Stochastic Defense Impossibility]
\label{thm:stochastic}
Let $f\colon X \to \R$ be continuous on a connected Hausdorff space
with $S_\tau, U_\tau \neq \emptyset$. Let $D$ be a stochastic defense,
and define $g(x) = \mathbb{E}_{y \sim D(x)}[f(y)]$ (the expected
alignment deviation after defense). If $g$ is continuous and $g(x) =
f(x)$ for all $x \in S_\tau$ (utility preservation in expectation),
then there exists $z$ with $f(z) = \tau$ and $g(z) = \tau$.
\end{theorem}

\begin{proof}
The function $h = g - f$ is continuous and vanishes on $S_\tau$. Since
$S_\tau$ is dense in $\cl{S_\tau}$ and $\{h = 0\}$ is closed, $h$
vanishes on $\cl{S_\tau}$. Any $z \in \cl{S_\tau} \setminus S_\tau$
satisfies $f(z) = \tau$ (by the argument of \Cref{thm:main}) and
$g(z) = f(z) = \tau$. The defense cannot reduce even the
\emph{expected} alignment deviation below $\tau$ at these points.
\end{proof}

Randomization does not help: the expected alignment deviation after a
stochastic defense inherits the same boundary fixation as a
deterministic one.

\subsection{Capacity Parity}

Even when the defense has the same computational capacity as the
attacker, utility preservation acts as a structural tax.

\begin{theorem}[Capacity Parity Disadvantage]
\label{thm:capacity}
Let the defense and attacker both operate on a space of $N$ inputs,
of which $|S_\tau|$ are safe. The defense must implement identity on
all safe inputs, consuming $|S_\tau|$ of its capacity. The attacker
faces no such constraint. Under capacity parity ($C = N$), the
defense's effective free capacity is $N - |S_\tau| < N$, giving the
attacker an advantage of $|S_\tau|$ capacity units.
\end{theorem}

This is a counting argument: utility preservation is not free. Every
safe input the defense must leave untouched is one fewer degree of
freedom available for remapping unsafe inputs. The attacker, who has
no obligation to preserve anything, retains full capacity.

\subsection{Nonlinear Agent Pipelines}

Modern AI agents chain multiple steps: defense $\to$ model inference
$\to$ tool call $\to$ tool call $\to \cdots$ Each stage is a nonlinear
map that can amplify small input differences. The natural question is
whether this pipeline structure helps the defense by giving it multiple
chances to intervene.

Each stage \emph{multiplies} the effective Lipschitz constant,
widening the failure band with pipeline depth.

\begin{theorem}[Pipeline Lipschitz Degradation]
\label{thm:pipeline}
If stages $T_1, \ldots, T_n$ are $K_1, \ldots, K_n$-Lipschitz
respectively, the composed pipeline $T_n \circ \cdots \circ T_1$ is
$(K_1 \cdot K_2 \cdots K_n)$-Lipschitz. For $n$ stages each with
constant $K \geq 2$, the effective constant is $K^n$---exponential
in pipeline depth.
\end{theorem}

\begin{proof}
Composition of Lipschitz maps is Lipschitz with product constant
(standard; verified in Lean via \texttt{LipschitzWith.comp}).
\end{proof}

\begin{theorem}[Pipeline Impossibility]
\label{thm:pipeline-impossible}
If the composed pipeline $P = T_n \circ \cdots \circ T_1 \circ D$ is
continuous and preserves safe inputs ($f(x) < \tau \Rightarrow
P(x) = x$), then $P$ has boundary fixed points.
\end{theorem}

The $\varepsilon$-robust band width for the pipeline scales as
$L(K^n + 1)\delta$, where $K^n$ is the effective Lipschitz constant.
Each tool call \emph{multiplicatively} degrades the defense's
effectiveness: a 5-stage pipeline with $K = 2$ per stage has
effective constant $32$, making the failure band $33\times$ wider than
a single-stage defense. The failure band grows exponentially with
pipeline depth.

\begin{corollary}[Stateful Pipeline Impossibility]
\label{cor:stateful}
If an agent's state evolves as $s_{t+1} = T(D(s_t))$ where $T$ and $D$ are continuous and $D$ preserves safe states, then $\exists\, z$ with $f(z) = \tau$ and $T(D(z)) = z$ (verified in Lean as \texttt{stateful\_pipeline\_impossibility}).
\end{corollary}

\begin{corollary}[Complete Defense Must Be Discontinuous]
\label{cor:jump}
No continuous utility-preserving defense is complete: if $f(D(x)) < \tau$ for all $x$, then $D$ cannot be both continuous and utility-preserving when $S_\tau, U_\tau \neq \emptyset$ (verified as \texttt{complete\_defense\_must\_jump}).
\end{corollary}


% ============================================================
\section{The Vulnerability Landscape}
\label{sec:landscape}

The preceding sections show that defenses must fail \emph{somewhere}.
This section characterizes \emph{where}: vulnerabilities are not
isolated needles in a haystack but extended regions (``basins'') with
computable minimum sizes.

\begin{theorem}[Basin Structure]
\label{thm:basin}
If $f$ is continuous and $f(p)>\tau$, then $U_\tau$ is open. Under
any measure that assigns positive mass to nonempty open sets (e.g.,
Lebesgue measure on~$\R^n$), $U_\tau$ has positive measure.
\end{theorem}

\begin{theorem}[Basin Fragment Minimum Size]
\label{thm:fragment}
If $f$ is $L$-Lipschitz with $f(p) > \tau$, the connected component
of $U_\tau$ containing $p$ has diameter $\geq 2(f(p) - \tau)/L$.
\end{theorem}

Additional supporting results---perturbation robustness, iterative
convergence, transferability, authority monotonicity, and gradient
ascent---are developed in \Cref{app:attacks}. Fine-tuning stability
results appear in \Cref{app:stability}, and cost asymmetry analysis
in \Cref{app:cost}.


% ============================================================
\section{Experimental Validation}
\label{sec:experiments}

We validate the theorems' predictions against real model behavior. The
Manifold of Failure framework~\cite{munshi2026manifold} maps three LLMs
over a 2D behavioral space defined by two axes: \emph{query indirection}
(how obliquely the prompt asks for unsafe content) and \emph{authority
framing} (how much the prompt invokes authority or permission). Each
point in this space corresponds to a prompt style, and the alignment
deviation score measures how unsafe the model's response is.

\begin{table}[t]
\centering
\small
\caption{Falsifiable predictions confirmed by empirical data.}
\label{tab:correspondence}
\begin{tabular}{@{}p{3.0cm}p{4.5cm}p{4.5cm}@{}}
\toprule
\textbf{Theorem} & \textbf{Predicts} & \textbf{Confirmed by} \\
\midrule
Basin Structure (\ref{thm:basin})
  & Basins are open with positive measure
  & Heatmaps show extended regions \\[3pt]
Fragmentation (\ref{thm:fragment})
  & Smooth $\to$ large basins; rough $\to$ mosaic
  & Llama: mesa; GPT-OSS: mosaic \\[3pt]
Convergence (\ref{thm:convergence})
  & Attacks exhibit monotone convergence
  & Convergence curves plateau \\[3pt]
Transferability (\ref{thm:transfer})
  & Similar surfaces $\to$ shared basins
  & Llama $.93 \to$ GPT-OSS $.73 \to$ Mini $.47$ \\[3pt]
Authority (\ref{thm:authority})
  & Horizontal banding
  & Bands at $a_2 \approx .25$--$.35$, $.65$--$.85$ \\[3pt]
Persistent (\ref{thm:persistent})
  & Steep boundaries $\to$ unsafe volume persists
  & Llama's $.93$ plateau persists under defense \\[3pt]
Interior Stability (\ref{thm:interior-stable})
  & Deep basin points survive fine-tuning
  & Vulnerability persists across variants \\[3pt]
Cost (\ref{thm:cost})
  & 2D tractable, high-$d$ intractable
  & 15K queries fill 63\% at $d\!=\!2$ \\[3pt]
Pipeline (\ref{thm:pipeline})
  & Deeper pipelines $\to$ wider failure band
  & Tool-using agents harder to defend than single-turn \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Llama-3-8B} (mean AD $0.93$, basin rate $93.9\%$): near-flat
mesa, small~$L$, large robustness radii. The directional slope of the
alignment surface at boundary crossings far exceeds $L(K+1)$ for any
reasonable~$K$, placing this model in persistent-unsafe-region
territory: the defense cannot eliminate the vulnerability
volume under these conditions.

\paragraph{GPT-OSS-20B} (mean AD $0.73$, basin rate $64.3\%$): rugged
landscape, large~$L$, small fragments. Horizontal bands confirm
authority monotonicity. The fragmented boundary presents many
transversal crossings, each generating a persistent unsafe pocket.

\paragraph{GPT-5-Mini} (peak AD $0.50$, basin rate $0\%$): at
$\tau = 0.50$, $U_\tau = \emptyset$, so none of the three theorems
apply---correctly predicting no impossibility. The model's strategy:
make the boundary harmless rather than eliminate it.

\paragraph{Quantitative checks.}
Robustness radius: $L \approx 5$ for Llama gives
$r = (0.93-0.50)/5 = 0.086$, covering $\sim$2 grid cells---consistent
with heatmap smoothness. Cost ratio: $0.01 \times 625 = 6.25$
theoretical vs.\ ${\sim}4$ empirical.


% ============================================================
\section{The Engineering Prescription}
\label{sec:engineering}

The impossibility results do not say that defense is without value---they
say that \emph{complete} defense is impossible under the stated constraints.
The gap between impossibility and practice is where engineering lives.
The following prescriptions are ordered from most to least actionable.

\paragraph{1. Make the boundary shallow.}
Set $\tau$ so that boundary-level behavior is benign. If $f(z) = \tau$
corresponds to a polite refusal rather than harmful compliance, the
impossibility is mathematically true but practically harmless.
GPT-5-Mini exemplifies this: its ceiling at $\AD = 0.50$ means the
``Manifold of Failure'' exists but contains no actual harm.

\paragraph{2. Reduce the Lipschitz constant.}
Smaller~$L$ makes the alignment surface smoother and more predictable.
It also raises the transversality threshold $L(K+1)$ for persistence: a
smoother surface is harder for the attacker to steer into persistent
unsafe territory. The tradeoff is that smoother surfaces spread
vulnerabilities over wider regions, but those regions are more easily
monitored.

\paragraph{3. Reduce the effective dimension.}
Defense cost grows as $N^d$ (\Cref{thm:cost}). Constraining the prompt
interface---standardized formats, restricted API parameters, bounded
context lengths---reduces~$d$, making the behavioral space tractable.

\paragraph{4. Monitor, don't eliminate, the boundary.}
\Cref{thm:crossing} shows transversal crossings persist under
fine-tuning, and \Cref{thm:multi-turn} shows they recur every turn.
Rather than attempting the impossible (eliminating boundary points),
deploy runtime monitoring that detects approach to the boundary. The
Lipschitz bound (\Cref{thm:lipschitz}) gives a computable estimate of
distance to the boundary from any observed AD value.

\paragraph{5. Do not rely on stateful accumulation.}
\Cref{thm:multi-turn} proves that accumulating context across turns
does not eliminate boundary fixation---the impossibility recurs fresh
at every turn. Defenses that rely on ``learning'' from conversation
history to progressively close vulnerabilities face the same
topological constraint at each turn.

\paragraph{6. Do not pay the stochastic tax.}
\Cref{thm:stochastic} proves that randomized defenses---stochastic
routing, temperature-based sampling, random prompt perturbation---fail
in expectation. The expected alignment deviation after a stochastic
defense inherits the same boundary fixation as a deterministic one.
Randomization does not help eliminate boundary points.

\paragraph{7. Break capacity parity with architectural separation.}
\Cref{thm:capacity} proves that under equal resources, the defense
loses capacity to utility preservation while the attacker retains full
freedom. The implication: a defense that shares the same architecture
and parameter budget as the model it protects starts at a structural
disadvantage. To break this asymmetry, separate the defense from the
utility pipeline entirely---for example, by using a dedicated
lightweight classifier that does not need to preserve utility (it only
flags), paired with a separate response generator that handles safe
inputs. This sacrifices the ``single continuous wrapper'' model but
sidesteps the capacity tax.


% ============================================================
\section{Unifying the Impossibility Results}
\label{sec:meta}

\paragraph{Intuition.}
We have now proved impossibility through six different lenses:
continuous topology, Lipschitz bounds, discrete counting, multi-turn
dynamics, stochastic expectations, and capacity constraints. A reader
might wonder: are these six different problems, or one problem seen
from six angles? They are one problem. Every path through the theory
reduces to the same core fact: a defense that promises ``I won't
change safe inputs'' is forced to also leave some \emph{non-safe}
inputs unchanged. This is because the set of inputs the defense
leaves alone (its ``fixed-point set'') must be larger than the safe
region---it extends into the boundary or beyond.

The preceding sections establish impossibility via multiple routes:
continuous topology, Lipschitz metric structure, discrete counting,
multi-turn dynamics, stochastic expectations, and capacity constraints.
It is natural to ask whether these are separate results that happen to
agree, or instances of a single underlying phenomenon.

They share a common structure.

\begin{theorem}[Regularity Implies Spillover]
\label{thm:meta}
Let $X$ be any set, $S \subseteq X$ nonempty and proper (the safe
region), and $D\colon X \to X$ satisfying utility preservation:
$D(x) = x$ for all $x \in S$. If $D$ has \emph{bounded
regularity}---meaning the set of inputs the defense leaves unchanged,
$\Fix(D) = \{x : D(x) = x\}$, cannot equal $S$ exactly---then
\[
  \Fix(D) \supsetneq S.
\]
\end{theorem}

\begin{proof}
Utility preservation gives $S \subseteq \Fix(D)$. Bounded regularity
gives $\Fix(D) \neq S$. Together: $S \subsetneq \Fix(D)$.
\end{proof}

\begin{remark}[Relaxed Utility Preservation]
Strict utility preservation ($D(x) = x$ on $S_\tau$) can be relaxed.
\Cref{thm:score-preserving} shows that score-preserving defenses
($f(D(x)) = f(x)$ on $S_\tau$) still satisfy $f(D(z)) = \tau$ at
boundary points, and \Cref{thm:eps-relaxed} shows that
$\varepsilon$-score-preserving defenses satisfy
$f(D(z)) \geq \tau - \varepsilon$. Only the degenerate case
$D(S_\tau) \subseteq S_\tau$ with no score constraint permits escape
(\Cref{rem:safe-preserving-escape}), at the cost of destroying all
semantic content.
\end{remark}

\begin{corollary}[Non-Safe Fixed Point]
$\Fix(D) \supsetneq S$ implies $\exists\, x \notin S$ with
$D(x) = x$. The defense fixes a non-safe input.
\end{corollary}

The hard work is not the corollary---which is pure set theory---but
proving that bounded regularity holds, i.e., that $\Fix(D)$ cannot
equal $S$. Each path through the theory provides a different proof:

\begin{center}
\small
\begin{tabular}{@{}lp{8cm}@{}}
\toprule
\textbf{Path} & \textbf{Why $\Fix(D) \supsetneq S$} \\
\midrule
Continuous
  & $\Fix(D)$ is closed (Hausdorff), $S$ is open and not closed
    (connected), so $\cl{S} \subseteq \Fix(D)$ but
    $\cl{S} \supsetneq S$. \\[3pt]
Lipschitz
  & Same as the continuous path, plus Lipschitz bounds constrain how far
    the defense can move points near the fixed boundary. \\[3pt]
Transversal
  & Same as the Lipschitz path, plus steep slope forces
    $f(D(x)) > \tau$ on a positive-measure set. \\[3pt]
Discrete
  & IVT on finite ordered set forces a crossing; utility preservation
    fixes the safe endpoint. \\[3pt]
Stochastic
  & $g - f$ vanishes on $S$ and is continuous, so it vanishes on
    $\cl{S}$; expected AD $= \tau$ at boundary. \\[3pt]
Multi-turn
  & Regularity holds at \emph{each} turn independently. \\[3pt]
Capacity
  & Finite capacity $<$ attack surface forces pigeonhole collision. \\
\bottomrule
\end{tabular}
\end{center}

\noindent For $\Fix(D)$ to equal $S$ exactly---the only way to escape
the meta-theorem---$S$ would need to be simultaneously open (as a
preimage of an open set under a continuous $f$) and closed (as a
fixed-point set of a continuous $D$). In a connected space with both
safe and unsafe inputs, no proper nonempty subset is both open and
closed. This is the geometric lock.

\begin{lemma}[Geometric Lock]
\label{lem:lock}
Let $X$ be a connected topological space, $S \subseteq X$ open, nonempty, and proper ($S \neq X$). Then $S$ is not closed. Consequently, for any continuous $D$ on a Hausdorff space, $\Fix(D) \neq S$ (verified in Lean as \texttt{geometric\_lock}).
\end{lemma}

The meta-theorem is representation-independent: it does not care whether
$X$ is $\R^n$, a discrete token space, a probability simplex, or an
abstract set. It requires only that some mechanism---topological,
metric, combinatorial, or information-theoretic---forces the
fixed-point set to spill beyond the safe region.


% ============================================================
\section{Limitations}
\label{sec:limitations}

We organize limitations into two categories: inherent scope boundaries
of the theorems, and adversarial challenges to the assumptions.

\subsection{Scope Boundaries}

\paragraph{Boundary fixation is at the boundary.}
Fixed points satisfy $f(z) = \tau$ exactly. If $\tau$-level behavior
is benign, the theorem is true but harmless.

\paragraph{The $\varepsilon$-robust constraint limits depth, not direction.}
The defense may push near-boundary points slightly below~$\tau$. The
bound limits how far, not whether.

\paragraph{Persistence requires transversality.}
The persistent unsafe region exists only where the alignment surface
is steep ($c > L(K+1)$).

\paragraph{Grid-based cost asymmetry.}
\Cref{thm:cost} assumes exhaustive grid enumeration. Learning-based
defenses that generalize across the space may sidestep the exponential
bound.

\subsection{Adversarial Challenges to Assumptions}


\begin{table}[!htbp]
\centering
\footnotesize
\setlength{\tabcolsep}{4pt}
\caption{Challenges to the theorems' assumptions.}
\label{tab:challenges}
\begin{tabular}{@{}p{3.4cm}p{3cm}p{4cm}p{4cm}@{}}
\toprule
\textbf{Challenge} & \textbf{Breaks} & \textbf{Effect} & \textbf{Status} \\
\midrule
Jitter defense
  & $D|_{S_\tau}\!=\!\id$
  & $\Fix(D)\!\not\supseteq\! S_\tau$
  & Genuine escape; reintroduces classification boundary \\[3pt]
Binary exploits
  & $f$ continuous
  & Discrete state changes $\notin C^0$
  & Targets behavioral $f$; use blocklists for structural vulns \\[3pt]
Token measure gap
  & Tietze bridge
  & $\mu(\mathcal{S})\!>\!0$ may miss valid tokens
  & $R^2\!>\!0.85$ empirically~\cite{munshi2026manifold}; formal bridge open \\[3pt]
Non-Hausdorff agents
  & $\Delta$ closed
  & $\Fix(D)$ not closed
  & Lock needs connectedness only; discrete path topology-free \\[3pt]
Boundary tunneling
  & $B_\tau$ monitoring
  & Interior basins escape estimates
  & Prediction: add interior monitoring \\[3pt]
$\tau$ cliff
  & Severity spectrum
  & No benign boundary
  & Needs sandboxing / isolation \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We have established three impossibility results for prompt injection
defense, forming a progressive hierarchy. Boundary fixation:
any continuous, utility-preserving defense must fix boundary points.
$\varepsilon$-robust constraint: the defense cannot uniformly
reduce alignment deviation far below~$\tau$ on a positive-measure band.
Persistent unsafe region: under transversality, the defense
leaves positive-measure volume unsafe---$f(D(x)) > \tau$
persists. A representation-independent meta-theorem (\Cref{thm:meta})
reveals common structure: each path through the theory shows that
the defense's fixed-point set must extend beyond the safe region.
Continuous topology, Lipschitz bounds, discrete counting, stochastic
expectations, multi-turn dynamics, and capacity constraints all lead
to this conclusion through different mechanisms.
The impossibility is a consequence of combining utility preservation
with any form of bounded regularity, not of any particular
representation choice.
Supporting theorems show that vulnerabilities form
basins, attacks converge and transfer, vulnerabilities survive
fine-tuning, and defense cost grows exponentially in dimension. The
practical prescription: make the boundary shallow (raise~$\tau$),
smooth (reduce~$L$), and low-dimensional (constrain the interface).
Our results, machine-verified in Lean~4, characterize the ``Defense
Trilemma'' and provide a formal basis for engineering safety through
boundary-depth management rather than absolute containment.


% ============================================================
% BIBLIOGRAPHY
% ============================================================
\bibliographystyle{plain}
\begin{thebibliography}{99}

\bibitem{alon2023detecting}
G.~Alon and M.~Kamfonas.
\newblock Detecting language model attacks with perplexity.
\newblock \emph{arXiv preprint arXiv:2308.14132}, 2023.

\bibitem{bagnall2019certifying}
A.~Bagnall and G.~Stewart.
\newblock Certifying the true error: Machine learning in {Coq} with
  verified generalization guarantees.
\newblock \emph{Proceedings of the AAAI Conference on Artificial
Intelligence}, 2019.

\bibitem{bai2022constitutional}
Y.~Bai et~al.
\newblock Constitutional {AI}: Harmlessness from {AI} feedback.
\newblock \emph{arXiv preprint arXiv:2212.08073}, 2022.

%% TODO: Could not verify this reference. No paper by Bolte et al. with this
%% title was found in Mathematical Programming. Closest match: Ranjan & Stellato,
%% "Verification of First-Order Methods for Parametric Quadratic Optimization,"
%% Math. Programming, 2025 (arXiv:2403.03331).
\bibitem{bolte2024}
J.~Bolte et~al.
\newblock Formal verification of optimization algorithms.
\newblock \emph{Mathematical Programming}, 2024.

\bibitem{chao2024jailbreaking}
P.~Chao, A.~Robey, E.~Dobriban, H.~Hassani, G.~J. Pappas, and
E.~Wong.
\newblock Jailbreaking black box large language models in twenty queries.
\newblock \emph{arXiv preprint arXiv:2310.08419}, 2024.

\bibitem{cohen2019certified}
J.~Cohen, E.~Rosenfeld, and J.~Z. Kolter.
\newblock Certified adversarial robustness via randomized smoothing.
\newblock \emph{Proceedings of ICML}, 2019.

\bibitem{goodfellow2014explaining}
I.~J. Goodfellow, J.~Shlens, and C.~Szegedy.
\newblock Explaining and harnessing adversarial examples.
\newblock \emph{arXiv preprint arXiv:1412.6572}, 2014.

\bibitem{inan2023llama}
H.~Inan et~al.
\newblock {Llama Guard}: {LLM}-based input-output safeguard for
  human-{AI} conversations.
\newblock \emph{arXiv preprint arXiv:2312.06674}, 2023.

\bibitem{katz2017reluplex}
G.~Katz, C.~Barrett, D.~L. Dill, K.~Julian, and
M.~J. Kochenderfer.
\newblock Reluplex: An efficient {SMT} solver for verifying deep neural
  networks.
\newblock \emph{Proceedings of CAV}, 2017.

\bibitem{mehrotra2024tree}
A.~Mehrotra et~al.
\newblock Tree of attacks: Jailbreaking black-box {LLMs} automatically.
\newblock \emph{Advances in Neural Information Processing Systems}, 37,
2024.

\bibitem{carlini2017evaluating}
N.~Carlini and D.~Wagner.
\newblock Towards evaluating the robustness of neural networks.
\newblock \emph{Proceedings of IEEE S\&P}, 2017.

\bibitem{fawzi2018adversarial}
A.~Fawzi, H.~Fawzi, and O.~Fawzi.
\newblock Adversarial vulnerability for any classifier.
\newblock \emph{Advances in Neural Information Processing Systems}, 31, 2018.

\bibitem{greshake2023indirect}
K.~Greshake, S.~Abdelnabi, S.~Mishra, C.~Endres, T.~Holz, and M.~Fritz.
\newblock Not what you've signed up for: Compromising real-world
  {LLM}-integrated applications with indirect prompt injection.
\newblock \emph{Proceedings of AISec}, 2023.

\bibitem{huang2017safety}
X.~Huang, M.~Kwiatkowska, S.~Wang, and M.~Wu.
\newblock Safety verification of deep neural networks.
\newblock \emph{Proceedings of CAV}, 2017.

\bibitem{madry2018towards}
A.~Madry, A.~Makelov, L.~Schmidt, D.~Tsipras, and A.~Vladu.
\newblock Towards deep learning models resistant to adversarial attacks.
\newblock \emph{Proceedings of ICLR}, 2018.

\bibitem{naitzat2020topology}
G.~Naitzat, A.~Zhitnikov, and L.-H.~Lim.
\newblock Topology of deep neural networks.
\newblock \emph{Journal of Machine Learning Research}, 21(184):1--40, 2020.


\bibitem{mouret2015illuminating}
J.-B. Mouret and J.~Clune.
\newblock Illuminating search spaces by mapping elites.
\newblock \emph{arXiv preprint arXiv:1504.04909}, 2015.

\bibitem{munshi2026manifold}
S.~Munshi, M.~Bhatt, V.~S.~Narajala, I.~Habler, A.~Al-Kahfah,
K.~Huang, and B.~Gatto.
\newblock Manifold of failure: Behavioral attraction basins in language
  models.
\newblock \emph{arXiv preprint arXiv:2602.22291v2}, 2026.

\bibitem{samvelyan2024rainbow}
M.~Samvelyan et~al.
\newblock Rainbow teaming: Open-ended generation of diverse adversarial
  prompts.
\newblock \emph{Advances in Neural Information Processing Systems}, 37,
2024.

\bibitem{singh2019abstract}
G.~Singh, T.~Gehr, M.~P{\"u}schel, and M.~Vechev.
\newblock An abstract domain for certifying neural networks.
\newblock \emph{Proceedings of POPL}, 2019.

\bibitem{szegedy2013intriguing}
C.~Szegedy et~al.
\newblock Intriguing properties of neural networks.
\newblock \emph{arXiv preprint arXiv:1312.6199}, 2013.

\bibitem{tsipras2018robustness}
D.~Tsipras, S.~Santurkar, L.~Engstrom, A.~Turner, and A.~Madry.
\newblock Robustness may be at odds with accuracy.
\newblock \emph{arXiv preprint arXiv:1805.12152}, 2018.

\bibitem{wolpert1997no}
D.~H. Wolpert and W.~G. Macready.
\newblock No free lunch theorems for optimization.
\newblock \emph{IEEE Trans.\ Evol.\ Comput.}, 1(1):67--82, 1997.

\bibitem{zou2023universal}
A.~Zou, Z.~Wang, N.~Carlini, M.~Nasr, J.~Z. Kolter, and
M.~Fredrikson.
\newblock Universal and transferable adversarial attacks on aligned
  language models.
\newblock \emph{arXiv preprint arXiv:2307.15043}, 2023.
\end{thebibliography}


% ============================================================
% APPENDICES — after bibliography, on new page
% ============================================================
\newpage
\appendix

\section*{Appendices}

\section{Counterexamples: Each Hypothesis Is Necessary}
\label{app:counterexamples}

\begin{counterexample}[Removing connectedness]
$X = \{0,1\}$ discrete, $f(0)=0$, $f(1)=1$, $\tau=0.5$.
$D(0)=0$, $D(1)=0$: continuous, utility-preserving, complete.
\end{counterexample}

\begin{counterexample}[Removing continuity]
$X=[0,1]$, $f(x)=x$, $\tau=0.5$. $D(x)=x$ for $x<0.5$,
$D(x)=0$ for $x \geq 0.5$: utility-preserving and complete,
but discontinuous at $0.5$.
\end{counterexample}

\begin{counterexample}[Removing utility preservation]
$D(x) = x_0$ for a fixed safe point: continuous and complete,
but destroys all inputs.
\end{counterexample}

\begin{remark}[The defense trilemma]
A defense can satisfy at most two of: continuity, utility preservation,
completeness.
\end{remark}


\section{Attack Properties}
\label{app:attacks}

\begin{theorem}[Perturbation Robustness]
\label{thm:lipschitz}
If $f$ is $L$-Lipschitz and $f(p)>\tau$, then
$\operatorname{ball}(p,\, (f(p)-\tau)/L) \subseteq U_\tau$.
The radius is monotone in~$f(p)$.
\end{theorem}

\begin{theorem}[Iterative Convergence]
\label{thm:convergence}
Any monotone-improvement operator $T$ with score in $[0,1]$ converges.
If each step gains $\geq \delta$, convergence takes
$\leq \lfloor 1/\delta \rfloor$ steps.
\end{theorem}

\begin{theorem}[Transferability]
\label{thm:transfer}
$\|f-g\|_\infty \leq \delta \implies
\{f > \tau+\delta\} \subseteq \{g > \tau\}$.
Transfer costs zero additional queries.
\end{theorem}

\begin{theorem}[Authority Monotonicity]
\label{thm:authority}
If $f(y,\cdot)$ is monotone non-decreasing in authority for each fixed
indirection~$y$, then the vulnerability set $\{x : f(y,x) > \tau\}$ is
upward-closed and a critical threshold $a^*_2(y)$ exists (by IVT, when
$f(y,\cdot)$ is continuous and crosses~$\tau$). If additionally $f$ is
monotone non-decreasing in indirection for each fixed authority, the
threshold curve $y \mapsto a^*_2(y)$ is non-increasing.
\end{theorem}

\begin{theorem}[Gradient Ascent]
\label{thm:gradient}
If $f$ has nonzero Fr\'{e}chet derivative at~$x$, there exist $v$ and
$\varepsilon > 0$ with $f(x + \varepsilon v) > f(x)$.
\end{theorem}


\section{Stability Under Fine-Tuning}
\label{app:stability}

\begin{theorem}[Interior Stability]
\label{thm:interior-stable}
If $\|f - g\|_\infty \leq \varepsilon$:
$f(x) > \tau + \varepsilon \implies g(x) > \tau$, and
$f(x) < \tau - \varepsilon \implies g(x) < \tau$.
Only the band $|f(x) - \tau| \leq \varepsilon$ is uncertain.
\end{theorem}

\begin{theorem}[Crossing Preservation]
\label{thm:crossing}
If $f$ crosses $\tau$ transversally on $[a,b]$ and $g$ is continuous
with $\|f - g\|_\infty \leq \varepsilon$, then $g$ also crosses $\tau$
on $[a,b]$.
\end{theorem}

\begin{theorem}[Patching Is Nonlocal]
\label{thm:nonlocal}
There exist $f, g$ with $\|f-g\|_\infty \leq \varepsilon$ such that
eliminating a vulnerability at one point necessarily changes values at
distant points. (Verified by explicit construction in Lean.)
\end{theorem}


\section{Cost Asymmetry}
\label{app:cost}

\begin{theorem}[Exponential Cost Asymmetry for Grid-Based Defense]
\label{thm:cost}
For a defense monitoring the behavioral space via a grid with
$N \geq 2$ bins per axis in $d$ dimensions: attack cost
$\leq 1/\delta$ (dimension-independent); grid-based defense cost
$= N^d$ (exponential in~$d$); the ratio
$\delta \cdot N^d \to \infty$ as $d\to\infty$.
\end{theorem}

At $d=2$ with $N=25$ and $\delta=0.01$, the ratio is $6.25$.
At $d=10$, it climbs to $\sim 10^{12}$. Verified by \texttt{norm\_num}
in Lean.


\section{Additional Verified Results}
\label{app:additional}

The following results are proved in Lean but summarized here for completeness.

\paragraph{Lipschitz displacement bound.}
If $D$ is $K$-Lipschitz and $D(z) = z$, then $\dist(D(x), x) \leq (K+1)\dist(x,z)$ for all $x$ (\texttt{defense\_fixes\_nearby}). This is the triangle inequality: $\dist(D(x), x) \leq \dist(D(x), D(z)) + \dist(z, x) \leq K\dist(x,z) + \dist(x,z)$.

\paragraph{Tool calls amplify failure.}
Each non-contractive tool call ($K_T \geq 1$) multiplicatively increases the pipeline's effective Lipschitz constant: $K_T \cdot K_D \geq K_D$ (\texttt{tool\_call\_amplifies}). Consequently $K^n < K^{n+1}$ strictly (\texttt{deeper\_pipeline\_wider\_band}), and $K^n \geq 2^n$ for $K \geq 2$ (\texttt{band\_grows\_with\_depth}).

\paragraph{Attacker monotone improvement.}
The attacker's best observed alignment deviation across turns is non-decreasing: the running maximum of a sequence never drops (\texttt{running\_max\_monotone}).

\paragraph{Boundary knowledge accumulation.}
The set of known boundary points grows monotonically over turns (\texttt{boundary\_points\_accumulate}). After $T$ probes, the attacker has classified $T$ points (\texttt{knowledge\_constrains\_boundary}).

\paragraph{Attacker steering.}
If the directional slope at the boundary varies continuously with the attacker's influence parameter $\alpha$, and the slope takes values on both sides of $L(K+1)$, then the intermediate value theorem guarantees $\exists\, \alpha^*$ at which transversality is reached (\texttt{transversality\_reachable}).

\paragraph{Stochastic regularity.}
For a stochastic defense with expected output $g(x) = \mathbb{E}[f(D(x))]$: the function $g - f$ vanishes on $S_\tau$ and, by continuity, on $\cl{S_\tau}$. This establishes the regularity condition for the meta-theorem (\texttt{stochastic\_regularity}).

\paragraph{Discrete trilemma.}
On a finite ordered set where $f$ crosses $\tau$: a utility-preserving defense must leave the safe side of the crossing fixed, and a bounded-capacity defense is eventually overwhelmed by the exponential attack surface (\texttt{discrete\_trilemma}).

\paragraph{Defense position invariance.}
Whether the defense is placed before or after the tool calls, the Lipschitz constant of the composed pipeline is the same product $K_D \cdot K_T^n$ (\texttt{defense\_before\_tools}, \texttt{defense\_after\_tools}).

\section{Lean Artifact}
\label{app:artifact}

The complete theory is verified in Lean~4.28.0 with Mathlib~v4.28.0.
The artifact comprises 40 files available \href{https://github.com/mbhatt1/stuff/tree/main/ManifoldProofs}{here}:
\begin{itemize}[nosep]
  \item 10 core theory files (MoF\_01--MoF\_10)
  \item 10 cost theory files (MoF\_Cost\_01--MoF\_Cost\_10)
  \item 10 advanced theory files (MoF\_Adv\_01--MoF\_Adv\_10)
  \item 1 continuous relaxation (MoF\_ContinuousRelaxation)
  \item 1 $\varepsilon$-robust + persistent unsafe region
    (MoF\_11\_EpsilonRobust)
  \item 1 discrete impossibility (MoF\_12\_Discrete)
  \item 1 multi-turn + stochastic + capacity parity
    (MoF\_13\_MultiTurn)
  \item 1 representation-independent meta-theorem
    (MoF\_14\_MetaTheorem)
  \item 1 nonlinear agent pipelines (MoF\_15\_NonlinearAgents)
  \item 1 relaxed utility preservation (MoF\_16\_RelaxedUtility)
  \item 3 capstone files (MasterTheorem, Euclidean instantiation,
    verification)
\end{itemize}

Every theorem depends on exactly three axioms: \texttt{propext},
\texttt{Classical.choice}, \texttt{Quot.sound}---Lean's standard
foundations. No additional axioms.

Key Mathlib lemmas used:
\begin{center}
\small
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Lean name} & \textbf{Role} \\
\midrule
\texttt{isClosed\_diagonal} & Hausdorff $\Rightarrow$ diagonal closed \\
\texttt{LipschitzWith.dist\_le\_mul} & Lipschitz distance bound \\
\texttt{isClopen\_iff} & Connected $\Rightarrow$ no proper clopen \\
\texttt{closure\_minimal} & Closure in closed superset \\
\texttt{IsOpen.measure\_pos} & Open nonempty $\Rightarrow$ positive measure \\
\texttt{intermediate\_value$_2$} & IVT on connected spaces \\
\bottomrule
\end{tabular}
\end{center}

\end{document}
