% =====================================================================
% neurips_compute_disclosure.tex
% Included from paper2_v3_submission.tex via \input{}.
% Standalone section: total compute, cost, and wall-clock used for the
% empirical results in the paper. Does not affect the theoretical claims
% (which are Lean-verified, no compute involved).
% =====================================================================

\section*{Compute Disclosure}
\label{sec:compute}

The theoretical contributions of this paper---Theorems~\ref{thm:main}
through~\ref{thm:stochastic} and their refinements---are
\emph{Lean~4 / Mathlib} machine-verified and required no empirical
compute beyond a standard \texttt{lake build} pass (minutes on a
consumer laptop). All empirical compute below supports only
\Cref{sec:experiments} and \Cref{app:exp-details}.

\paragraph{OpenAI API usage.}
Behavioral-manifold construction and validator runs were executed
against the public OpenAI API. The aggregate API spend across all live
runs referenced in the main paper and appendix---spanning the two
canonical $25\times25$ MAP-Elites grids
(\texttt{gpt-5-mini-2025-08-07} and \texttt{gpt-3.5-turbo-0125}), the
saturated 400-evaluation \texttt{gpt-3.5-turbo-0125} archive
($\sim$5\,200 API calls, wall-clock $\sim$130 minutes, cost
$\sim$\$15), the seed-replication run (wall-clock $\sim$119 minutes),
two retargeted re-evaluation surfaces
(\texttt{gpt-4o-mini-2024-07-18} and \texttt{gpt-4o-2024-08-06}), a
66-prompt paraphrase pass with \texttt{gpt-4o-mini} at $T\!=\!0$, and
three judge-robustness and PAIR-baseline runs---totals approximately
\textbf{\$60--\$80 USD} at list prices. Individual run wall-clocks
recorded in \texttt{final\_statistics.json} files for the two largest
runs are \texttt{7\,814\,s} (saturated) and \texttt{7\,139\,s}
(replicate); total API wall-clock across all runs is under four
hours.

\paragraph{Local (Ollama) inference.}
The LlamaGuard-3-1B demonstration (\Cref{tab:llamaguard-demo}) was run
locally via Ollama on an Apple-silicon laptop (M-series, on-device
Metal acceleration). No external compute, no GPU cluster. The full
82-prompt pass completes in under ten minutes of wall-clock.

\paragraph{Formal verification.}
Lean~4.28.0 + Mathlib~v4.28.0 compiles the 46-file \texttt{ManifoldProofs}
tree in approximately 20 minutes on the same consumer laptop. No GPU
is used. Verification reproducibility is a function of Lean+Mathlib
version pinning and is independent of API availability.

\paragraph{Attribution to paper sections.}
The \$15 saturated-archive budget supports
\Cref{sec:live-validation,sec:gp-instance} and the entire
continuous-defense sweep (\Cref{app:continuous-sweeps}); the
remaining $\sim$\$45--\$65 is divided across the retargeting sweep
(\Cref{tab:three-target}), the judge-robustness deltas
(\Cref{app:judge-robustness}), seed replication
(\Cref{tab:seed-replication}), and the PAIR baseline
(\Cref{app:pair-baseline}). The forced-collapse demonstration
(\Cref{tab:forced-collapse,tab:paraphrase-unsafe}) uses the 66
paraphrases from the saturated archive and a single extra scoring
pass; marginal cost below \$2.

\paragraph{Reproducibility envelope.}
All budgets quoted are order-of-magnitude; re-running the pipeline
from a fresh checkout should land within a factor of $2$ on cost and
wall-clock, modulated by OpenAI list-price changes and model-server
latency. No experiment in this paper required specialized hardware,
exceeded a single laptop's memory footprint, or relied on
non-public data. The anonymized artifact repository contains the
exact run scripts, the raw \texttt{live\_runs/} metadata (including
\texttt{final\_statistics.json} for each run), and the
reconstruction paths from \texttt{live\_runs/} to every table and
figure in the paper.
